Appendix to "New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-02-06 supplementary/index-def.qmd
update index definition

# The `MIC` and `TIC` index

Both `MIC` and `TIC` are information-based indexes derived from mutual information. The indexes are defined based on a partition, $g(k, l)$, of the space $(y_1, y_2)$ into $k \times l$ rectangles, For example, $g(2, 3)$ divides the data space into 2 rectangles in the $y_2$ direction and 3 rectangles in the $y_1$ direction. Let $G$ denotes all the possible partition.

MIC finds the **maximum** mutual information over G where $k$ and $l$ are bounded by the grid size: $k \times l < B(n) = n^{\alpha}$. $alpha = 0.3$ is used in our simulation.

$$I_{MIC} = \max_{g \in G} \frac{I(y_1, y_2 | g)}{\log(\min(k^*, l^*))}  = \max_{g \in G} \frac{\sum_{y1} \sum_{y2} P(y_1, y_2) \log \frac{P(y_1)}{P(y_1) P(y_2)}}{\log(\min(k^*, l^*))}$$

where $k^*$ and $l^*$ are the number of rectangles in the optimal partition $g$.

TIC calculates the **sum** of mutual information over G

$$I_{TIC} = \sum_{g \in G} \frac{I(y_1, y_2 | g)}{\log(\min(k^*, l^*))}  = \sum_{g \in G} \frac{\sum_{y1} \sum_{y2} P(y_1, y_2) \log \frac{P(y_1)}{P(y_1) P(y_2)}}{\log(\min(k^*, l^*))}$$

# The `loess` and `splines` index

The `loess` and `splines` indexes detect non-linear structure in the projection, as captured by their respective model. These indexes are computed by regressing both axes against each other, and find the maximum variance from the residual that can be explained by the model. Let $e^{\text{model}}_{y_1 \sim y_2}$ denote the residual from the loess/splines model $y_1 \sim y_2$, and $e^{\text{model}}_{y_2 \sim y_1}$ denote the residual from the loess/spliens model $y_2 \sim y_1$. The two indexes are calculated as

 $$I_{\text{loess}}(A) = \max \left(1 - \frac{var(e^{\text{loess}}_{y_1 \sim y_2})}{var(y_1)}, 1 - \frac{var(e^{\text{loess}}_{y_2 \sim y_1})}{var(y_2)}\right)$$

$$I_{\text{spline}}(A) = \max \left(1 - \frac{var(e^{\text{spline}}_{y_1 \sim y_2})}{var(y_1)}, 1 - \frac{var(e^{\text{spline}}_{y_2 \sim y_1})}{var(y_2)}\right)$$

>>>>

# The MIC and TIC index

Maximal Information Coefficient (MIC) and Total Information Coefficient (TIC) [@reshef2016measuring] are information-based indexes derived from mutual information. These indexes calculate the mutual information after partitioning the space $(y_1, y_2)$ into $k \times l$ rectangles, with a constraint on the grid size to limit complexity: $k \times l < B(n) = n^{\alpha}$, where $n$ is the number of observations in the data. A commonly used value for $\alpha$ in the previous studies is 0.3 to balance flexiblity and overfitting.

The MIC index finds the **maximum** mutual information over possible grids configeration that satisfy the constraint, and then normalized by the grid size: $\log(\min(k^*, l^*))$:

$$I_{MIC} = \frac{I(y_1, y_2 | k^*, l^*)}{\log(\min(k^*, l^*))}$$

where $(k^*, l^*)$ is the grid configuration that maximizes the mutual information $I(y_1, y_2 | k, l)$ under the constraint $k \times l < B(n) = n^{0.3}$.

TIC calculates the **sum** of the normalized mutual information for all grid configurations under the size constraint:

$$I_{TIC} = \sum_{k \times l < n^{0.3}} \frac{I(y_1, y_2 | k, l)}{\log(\min(k, l))}$$

# The loess and splines index

The loess and splines indexes detect non-linear structure in the projection by capturing the variance explained by the loess/ spline model. To account for the orientation of the projection, the model is run for both $y_1 \sim y_2$ and $y_2 \sim y_1$ and the larger of the two is used as the index value. Let $e^{\text{model}}_{y_1 \sim y_2}$ and $e^{\text{model}}_{y_2 \sim y_1}$ denote the residual from the loess/splines model $y_1 \sim y_2$, and $y_2 \sim y_1$. The two indexes are calculated as:

 $$I_{\text{loess}} = \max \left(1 - \frac{var(e^{\text{loess}}_{y_1 \sim y_2})}{var(y_1)}, 1 - \frac{var(e^{\text{loess}}_{y_2 \sim y_1})}{var(y_2)}\right)$$

$$I_{\text{spline}} = \max \left(1 - \frac{var(e^{\text{spline}}_{y_1 \sim y_2})}{var(y_1)}, 1 - \frac{var(e^{\text{spline}}_{y_2 \sim y_1})}{var(y_2)}\right)$$
